Parametricity Consequences of GADTs Representations

### 2021-02-13 20-gadts/GADT-story.tex
work on gadt-story

GADTs are traditionally represented via Church encodings. In this
view, GADTs are completely determined by their syntax. For example,
the GADT $\mathsf{G}$ defined by
\[\begin{array}{l}
\mathsf{data\,G : * \to *}\\
\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
\end{array}\]
{\color{blue} Use $\top$ instead of $1$ throughout?}

@@

List types are the quintessential examples of ADTs. The shape of the
container underlying the list type
\[\mathsf{List\,a} = \mathsf{Nil} \,|\, \mathsf{Cons\,a\,(List\,a)}\]
%of lists of elements of type $\mathsf{a}$
is determined by the types of its two constructors $\mathsf{Nil\, ::\,
  List\, a}$ and $\mathsf{Cons \,::\,a \to \List\,a\to
  List\,a}$. These constructors specify that the data in a list of
type $\mathsf{List\,a}$ are arranged linearly. The shape underlying
the type $\mathsf{List\,a}$ is therefore given by the set $\nat$ of
natural numbers, and the positions in a structure of this type are
given by specific natural numbers. Since the argument to every
occurrence of the type constructor $\mathsf{List}$ in the right-hand
side of above definition is the same as on its left-hand side, the
type $\mathsf{List\,a}$ enforces the invariant that all of the data in
a structure of this type have the same type $\mathsf{a}$. In a similar
way, the tree type
\[\mathsf{Tree\, a} = \mathsf{Leaf\,a}
\,|\,\mathsf{Node\,(Tree\,a)\,a\,(Tree\,a)}\] of binary trees
has as its underlying shape the type of binary trees of units, and the
positions in a structure of this type are given by sequences of L (for
``left'') and R (for ``right'') navigating a path
%to either a node or a leaf in
through the structure. The type $\mathsf{Tree\, a}$ enforces the
invariant that all of the data at the nodes and leaves in a structure
of this type have the same type $\mathsf{a}$.

@@

These fixpoint equations are sensible at the level of types. But to
ensure that the syntactic fixpoint actually denotes a semantic object
computed as a real-live fixpoint, the semantic fixpoint calculation
must converge. This means that the fixpoint being taken must be of
{\em functors} on the category interpreting types~\cite{tfca}, rather
than mere functions between the sets interpreting them. Said
differently, functions between sets interpreting types must have
corresponding {\em functorial actions} on functions between
interpretations of types. Reflecting this requirement back into syntax
gives that the body of a primitive data type representation of an ADT
must support a $\mathsf{map}$ function, i.e., must be an instance of
Haskell's $\mathsf{Functor}$ class.

@@

\subsection{GADTs}\label{sec:gadts}

GADTs and their uses and ubiquity in modern computing. Note in
particular that GADTs are purposely defined not to behave
type-uniformly. We will explore the consequences of this in
Section~\ref{sec:GADT-semantics}.

IMPORT ENRICO

%Seeing even ADTs as fixpoints of higher-order functors opens the door
%to GADTs...

For ADTs, shape and data are independent. Choose the shape first and
then fill it. GADTs are not like ADTs: the shape of the ``container''
depends on the contained data, so that it's not possible to modify the
data while retaining the shape because shape and data are not
independent.  So if we try to add GADTs, then we have to account for
the dependency between shape and data. The relationship between (how)
the data (is constructed) and the shape is left out.  But the data in
a GADT shouldn't disappear just because you map a function over it!
This is odd! So in what sense to GADTs generalize ADTs? In what sense
are they even data types?

The representation choices are similar for GADTs. The difference is
that the two are not isomorphic.

Once we see that we are interested in fixpoints of higher-order
functors, we begin to relax the uniformity. This has naturally led to
consideration of nested types such as... and GADTs such as...

@@

>>>>

GADTs are traditionally represented via Church encodings. In this
view, GADTs are completely determined by their syntax. For example,
the GADT $\mathsf{G}$ defined by
\[\begin{array}{l}
\mathsf{data\,G : * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
\end{array}\]
{\color{blue} Use $\top$ instead of $1$ throughout?}

@@

List types are the quintessential examples of ADTs. The shape of the
container underlying the list type
\[\mathsf{List\,a} = \mathsf{Nil} \,|\, \mathsf{Cons\,a\,(List\,a)}\]
%of lists of elements of type $\mathsf{a}$
is determined by the types of its two constructors $\mathsf{Nil\, ::\,
  List\, a}$ and $\mathsf{Cons \,::\,a \to \List\,a\to
  List\,a}$. These constructors specify that the data in a list of
type $\mathsf{List\,a}$ are arranged linearly. The shape underlying
the type $\mathsf{List\,a}$ is therefore given by the set $\nat$ of
natural numbers, and the positions in a structure of this type are
given by specific natural numbers. Since the type argument to every
occurrence of the type constructor $\mathsf{List}$ in the right-hand
side of above definition is the same the type instance being defined
on its left-hand side, the type $\mathsf{List\,a}$ enforces the
invariant that all of the data in a structure of this type have the
same type $\mathsf{a}$. In a similar way, the tree type
\[\mathsf{Tree\, a} = \mathsf{Leaf\,a}
\,|\,\mathsf{Node\,(Tree\,a)\,a\,(Tree\,a)}\] of binary trees
has as its underlying shape the type of binary trees of units, and the
positions in a structure of this type are given by sequences of L (for
``left'') and R (for ``right'') navigating a path
%to either a node or a leaf in
through the structure. The type $\mathsf{Tree\, a}$ enforces the
invariant that all of the data at the nodes and leaves in a structure
of this type have the same type $\mathsf{a}$.

@@

These fixpoint equations are sensible at the level of types. But to
ensure that the syntactic fixpoint actually denotes a semantic object
computed as a real-live fixpoint, the semantic fixpoint calculation
must converge. This means that the fixpoint being taken must be of
{\em functors} on the category interpreting types~\cite{tfca}, rather
than of mere functions between the sets interpreting them. Said
differently, functions between sets interpreting types must have
corresponding {\em functorial actions} on functions between
interpretations of types. Reflecting this requirement back into syntax
gives that the body of a primitive data type representation of an ADT
must support a $\mathsf{map}$ function, i.e., must be an instance of
Haskell's $\mathsf{Functor}$ class.

@@

\subsection{GADT Syntax}\label{sec:gadts}

{\color{blue} polynomial} GADTs relax the restriction on the type
instances appearing in a data type definition. The special form of
GADTs known as {\em nested types} allow the data constructors of a
GADT to take as arguments data whose types involve type instances of
the GADT other than the one being defined. For example, the 
definition 
\[\mathsf{PTree \, a = PLeaf \,a \,|\, PNode \,(PTree\,(a \times
  a))}\] of the nested type $\mathsf{PTree\,a}$ of perfect trees
introduces the data constructor(s $\mathsf{PLeaf :: a \to PTree\,a}$
and) $\mathsf{PNode :: PTree \,(a\times a) \to PTree\,a}$. It enforces
not only the invariant that all of the data in a structure of type
$\mathsf{PTree\,a}$ is of the same type $\mathsf{a}$, but also the
invariant that all perfect trees have lengths that are powers of
2. More general GADTs allow their constructors both to take as
arguments {\em and return as results} data whose types involve type
instances of the GADT other than the one being defined. An example is
the following GADT, given in Haskell syntax:
\[\begin{array}{l}
\mathsf{data\, Seq : * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Const :\, a \to Seq\,a}\\
\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, :\, Seq \,a \to Seq\,b \to
  Seq\,(a \times b)}\\ 
\end{array}\]
Since the return type of the data constructor $\mathsf{Pair}$ is not
simply a variable, $\mathsf{Seq}$ is not a nested type.

By contrast with the ADT $\mathsf{List\, a}$, where the type parameter
$\mathsf{a}$ is integral to the type being defined, the type parameter
$\mathsf{a}$ in $\mathsf{PTree\,a}$ and $\mathsf{Seq\,a}$ appears as a
``dummy'' parameter used only to give the kind \verb|* -> *| of the
type constructors $\mathsf{PTree}$ and $\mathsf{Seq}$. This is
explicitly captured by the above GADT syntax. A GADT --- even a nested
type --- thus does not define a {\em family of inductive types}, one
for each type argument, like an ADT does, but instead defines an
entire family of types that must be constructed simultaneously. That
is, a GADT defines an {\em inductive family of types}.

\subsubsection{Church Encodings of GADTs}

The syntax of GADTs is consonant with programmers' frequent use of
them in situations where type uniformity is specifically to be
eschewed.  {\color{blue} Typical applications of GADTs include generic
  programming, modeling programming languages via higher-order
  abstract syntax, maintaining invariants in data structures, and
  expressing constraints in embedded domain-specific languages. GADTs
  have also been used to implement tagless
  interpreters~\cite{pl04,pr06,pvww06} by trading the definition of a
  universal value domain for a direct specification of the property of
  being a value. Other applications are described in,
  e.g.,~\cite{js15,rou06,MORE,UPDATED}. OTHER SITUATIONS NEEDING
  TYPE-BASED DISPATCH.}

GADTs are useful in such situations precisely because generalizing the
type arguments in the instances of a GADT appearing as the return
types of its data constructors entails a strong connection between a
GADTs shape and the data it contains.  With ADTs, we first choose the
shape of the container and then fill that container with data of
whatever type we like. Critically, the choice of shape is independent
of the data to be stored. With GADTs, however, the shape of the
``container'' may actually {\em depend} on (the type of) the data to
be contained. On other words, the shape and the data are not
independent! As a result, it may not always be possible to modify a
GADT's data while retaining its shape because changing (the type of)
the data may actually change the shape.

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, our
GADT calculus must support an equality type $\mathsf{Eq}$ that is
inhabited when its two type arguments are the same and is uninhabited
otherwise, i.e., the syntactic reflection of semantic equality. Such a
type can either be taken as primitive or
%, in a higher-kinded calculus such as $F\omega$,
can be represented by its own Church encoding
\[\mathsf{Eq\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
  f\,a\,a) \to f\,t_1\,t_2}\] 
Using $\mathsf{Eq}$ the GADT $\mathsf{Seq}$, e.g., can be written as 
{\color{red} START HERE}
\[ {\color{blue} WHAT GOES HERE??} \]

Note that Seq has no syntactic map (because Pair doesn't), i.e., is
not an instance of the Functor class, and is not required to in this
representation.

\subsubsection{GADTs as Primitive Data Types}

Moreover, because its type argument varies in the instances of
$\mathsf{Seq}$ appearing in the types of its data constructors,
$\mathsf{Seq}$ cannot be characterized by any first-order fixpoint
equation. It can, be viewed as a solution to the higher-order fixpoint
equation
\[\mathsf{H\,f\,a = a \,+\,Lan_{\lambda a\,b.\,a \times b}\,
  \lambda a\,b.\,f\,a \times f\,b}\] where $\mathsf{Lan_k\, g}$ is the
syntactic reflection of the {\em left Kan extension} $\Lan_K \,G$ of
the functor $G$ interpreting $\mathsf{g}$ along the functor $K$
interpreting $\mathsf{k}$~\cite{mac71}.

Other GADTs
are similarly characterizable by higher-order functors.  The
difference between ADTs and GADTs can thus be seen as a difference in
{\em orders} of their underlying functors. {\color{blue} The role of
  higher-order functors here is essential.}

{\color{blue} GADTs make essential use of higher-order functors.}

Need syntax functors (Lan)
The representation choices are similar for GADTs. The difference is
that the two are not isomorphic.

@@

But the data in a GADT shouldn't disappear just because you map a
function over it!  This is odd! So in what sense to GADTs generalize
ADTs? In what sense are they even data types?

The semantics of GADTs is not so clear.
